---
title: "Group 10 - Credit Scoring"
author: "Timmy Tschannen, Sarah Lea Schürch, Keijo Alexander Nierula, Roman Krass"
date: "2023-11-07"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE, echo=FALSE}
# Getting started by changing the default output of echo to TRUE for the current document 
knitr::opts_chunk$set(echo = TRUE)

# Create a list of packages to install and load into the work space
libraries = c("readr", "tidyverse", "dlookr", "ggpubr")
 
# Install packages from the predefined libraries list
lapply(libraries, function(x) if (!(x %in% installed.packages())) {
  install.packages(x)
})

# Load libraries
lapply(libraries, library, quietly = TRUE, character.only = TRUE)

# Remove current environment 
rm(list=ls())
```

```{r graph_setup, results="hide", message=FALSE}

# Define colors to create a general style 
custom_fill <- "#a8d1d1"
custom_alpha <- 0.8
custom_color <- "darkgrey"


```

# Project Setup {.unnumbered}
```{r }
# Set seed that the following results will stay the same when using the same seed
set.seed(7)

# Importing data
# Make sure, the loan_sample_10.csv-file is located in the same directory as this RMD-file
loanData <- read_csv("loan_sample_10.csv")

# Store loaded loan data in variable called "data"
data <- loanData

# Checking data for unusual things at the beginning or ending of the data set
head(data)
tail(data)
```
The `head()` and `tail()` functions show, that there is no unusual observation and the data set looks complete and has no emtpy rows at the end.

# Checking data for NA values {.unnumbered}
```{r}
# Check for missing values in each column of the data set and store the result in the "has_NAs" data frame
has_NAs = as.data.frame(apply(data, 2, function(x) any(is.na(x))))

# Map matching column names for a more readable output
colnames(has_NAs) = c("has_NA")

# Print the has_NAs data frame to check if there are any NA values in our data set
has_NAs
```
We also found out, that none of the variables has any `NA` values.

---

# Exercise 1

## Describe the data. Specifically:

### Check and report the structure of the data set.
```{r}
str(data)
```

### How many numeric and how many categorical variables are included in the data? What categorical variable has the most levels in it?
```{r}
str(data)
```
Reporting the structure of the data set shows us, that we have 5 character columns and 12 number columns.

### Summarize the variables. Discuss the summary statistics obtained.
```{r}
summary(data)
```

### Check the levels of the target variable by choosing the appropriate visualization. Is the target variable balanced?
```{r}
str(data)
```
While checking the summary, we can observe, that there might be some values with insane outliers. For example `annual_inc`, `revol_bal` or `tot_cur_bal`. These variables all have a small minimum compared to the median and the max value. This could mean that either the data is very wide spread or there are a few outliers that have an impact on the median and the mean.

### Check the distribution of the numeric variables in the data set (include different visual representations).
```{r}
#store all numeric data in new data set
data_numeric = data %>% select_if(is.numeric)

# make a loop to iterate through all numeric variables
for (i in 1:length(data_numeric[,])) {
  # get the column names to create the titles for the plots
  col_names = names(data_numeric)
  col_name_by_index =  col_names[i]
  
  boxplot <- ggplot(data,
       aes(x = `data_numeric[,i]`)) +
  geom_boxplot(
    outlier.color = "#006769",
    outlier.size  = 1,
    fill = "#006769",
    color = "black",
    alpha = .8) +
  labs(title = "Bitcoin box plot by coinbase exchange", x="Bitcoin price in dollars")
  
  print(ggarrange(boxplot))
}
```

## blabla
for (col in names(data_numeric)) {
  ggplot(data, aes_string(x = "", y = col)) +
    geom_boxplot() +
    labs(title = paste("Boxplot of", col))
}

